### 2021-08-12 content/80.formal-definition.tex
data commons figure as svg
Synchronized with Google Docs version 2021-08-12 to overleaf
.. and a couple more edits

https://docs.google.com/document/d/11toi7xIDiDDeOq4fA_pr1GUDf0julX0Q/edit?usp=sharing&ouid=108141746798163896498&rtpof=true&sd=true

% Because this file is added with \include, do not use
% Markdown setions within.

@@

\allowdisplaybreaks
\begin{eqnarray*}
&& ROCrate(\texttt{<http://example.com/ro/123/>}) \\
&& name(\texttt{<http://example.com/ro/123/>}, \\
&& \ \ \ \ \ \text{“Data files associated with the manuscript:Effects of …”}) \\
&& description(\texttt{<http://example.com/ro/123/}, \\
&& \ \ \ \ \ \text{“Palliative care planning for nursing home residents …”}) \\
&& license(\texttt{<http://example.com/ro/123/>}, \\ 
&& \ \ \ \ \ \texttt{<https://creativecommons.org/licenses/by-nc-sa/3.0/au/>}) \\ 
&& datePublished(\texttt{<http://example.com/ro/123/>}, \text{“2017-02-23”}) \\ 
&& hasPart(\texttt{<http://example.com/ro/123/>}, 
        \texttt{<http://example.com/ro/123/file.txt>}) \\
&& hasPart(\texttt{<http://example.com/ro/123/>}, 
        \texttt{<http://example.com/ro/123/interviews/>}) \\
\\
&& ContextualEntity(\texttt{<https://creativecommons.org/licenses/by-nc-sa/3.0/au/>}) \\
&& name(\texttt{<https://creativecommons.org/licenses/by-nc-sa/3.0/au/>},  \\
&& \ \ \ \ \  \text{“Attribution-NonCommercial-ShareAlike 3.0 Australia (CC BY-NC-SA 3.0 AU)”}) \\
\\
&& File(\texttt{<http://example.com/ro/123/survey.csv>}) \\
&& name(\texttt{<http://example.com/ro/123/survey.csv>}, 
        \text{“Survey of care providers”}) \\
\\
&& Dataset(\texttt{<http://example.com/ro/123/interviews/>}) \\
&& name(\texttt{<http://example.com/ro/123/interviews/>},  \\
&& \ \ \ \ \  \text{“Audio recordings of care provider interviews”}) 
\end{eqnarray*}

In reality many additional attributes from schema.org types like \url{http://schema.org/Dataset} and \url{http://schema.org/CreativeWork} would be used to further describe the RO-Crate and its entities, but as these are optional they do not form part of this formalisation.

@@

Note that in the JSON-LD serialization of RO-Crate, the expression of $Class$ and $Property$ is typically indirect: The JSON-LD \texttt{@context} maps to schema.org IRIs, which, when resolved as Linked Data, embed their formal definition as RDFa.

@@

An important RO-Crate principle is that of being \textbf{self-describing} Therefore the serialisation of the RO-Crate into a file should also describe itself in a \href{https://www.researchobject.org/ro-crate/1.1/root-data-entity.html#ro-crate-metadata-file-descriptor}{Metadata File Descriptor}, indicating it is $about$ (describing) the RO-Crate root data entity, and that it $conformsTo$ a particular version of the RO-Crate specification:

@@

Note that although the metadata file necessarily is an \emph{information resource} written to disk or served over the network (e.g. as JSON-LD), it is not considered to be a contained \emph{part} of the RO-Crate in the form of a \emph{data entity}, rather it is described only as a \emph{contextual entity}.

While in the conceptual model the \emph{RO-Crate Metadata File} can be seen as the top-level node that describes the \emph{RO-Crate Root}, in the formal model (and the JSON-LD format) the metadata file descriptor is an additional contextual entity and not affecting the depth-limit of the RO-Crate.

@@

This exposes the first order logic domain of discourse of IRIs, with rational numbers and strings as their corresponding JSON-LD representation. These production rules first grow the graph of $R$ by adding a transitive rule – anything described in $R$ which is related to $o$, means that $o$ is also mentioned by the $ROCrate(R)$. For simplicity this rule is one-way; in practice the graph can also contain free-standing contextual entities that have outgoing relations to data- and contextual entities.

>>>>

% Because this file is added with \include, do not use
% Markdown setions within.
% 
% Edit here before syncing to 80.formal-definition.md for HTML version

@@

\allowdisplaybreaks
\begin{eqnarray*}
&& ROCrate(\texttt{<http://example.com/ro/123/>}) \\
&& name(\texttt{<http://example.com/ro/123/>}, \\
&& \ \ \ \ \ \text{“Data files associated with the manuscript:Effects of …”}) \\
&& description(\texttt{<http://example.com/ro/123/}, \\
&& \ \ \ \ \ \text{“Palliative care planning for nursing home residents …”}) \\
&& license(\texttt{<http://example.com/ro/123/>}, \\ 
&& \ \ \ \ \ \texttt{<https://spdx.org/licenses/CC-BY-4.0>}) \\ 
&& datePublished(\texttt{<http://example.com/ro/123/>}, \text{“2017-02-23”}) \\ 
&& hasPart(\texttt{<http://example.com/ro/123/>}, 
        \texttt{<http://example.com/ro/123/file.txt>}) \\
&& hasPart(\texttt{<http://example.com/ro/123/>}, 
        \texttt{<http://example.com/ro/123/interviews/>}) \\
\\
&& ContextualEntity(\texttt{<https://spdx.org/licenses/CC-BY-4.0>}) \\
&& name(\texttt{<https://spdx.org/licenses/CC-BY-4.0>},  \\
&& \ \ \ \ \  \text{Creative Commons Attribution 4.0”}) \\
\\
&& ContextualEntity(\texttt{<https://spdx.org/licenses/CC-BY-NC-4.0>}) \\
&& name(\texttt{<https://spdx.org/licenses/CC-BY-NC-4.0>},  \\
&& \ \ \ \ \  \text{Creative Commons Attribution Non Commercial 4.0”}) \\
\\
&& File(\texttt{<http://example.com/ro/123/survey.csv>}) \\
&& name(\texttt{<http://example.com/ro/123/survey.csv>}, 
        \text{“Survey of care providers”}) \\
\\
&& Dataset(\texttt{<http://example.com/ro/123/interviews/>}) \\
&& name(\texttt{<http://example.com/ro/123/interviews/>},  \\
&& \ \ \ \ \  \text{“Audio recordings of care provider interviews”}) \\
&& license(\texttt{<http://example.com/ro/123/interviews/>}, \\ 
&& \ \ \ \ \ \texttt{<https://spdx.org/licenses/CC-BY-NC-4.0>})
\end{eqnarray*}

Notable from this triple-like formalization is that a RO-Crate $R$ is fully represented as a tree at depth 2 helped by the use of $𝕀𝕣𝕚$ nodes. For instance the aggregation from the root entity $hasPart(…interviews/>)$ is at same level as the data entity’s property $license(…CC-BY-NC-4.0>)$ and that contextual entity’s attribute name $(…Non Commercial 4.0”)$. As shown in section RO-Crate JSON-LD, the RO-Crate Metadata File serialization is an equivalent shallow tree, although at depth 3 to cater for the JSON-LD preamble of \texttt{"@context"} and \texttt{"@graph"}.

In reality many additional attributes and contextual types from schema.org types like \url{http://schema.org/affiliation} and \url{http://schema.org/Organization} would be used to further describe the RO-Crate and its entities, but as these are optional (_SHOULD_ requirements) they do not form part of this formalization

@@

Note that in the JSON-LD serialization of RO-Crate, the expression of $Class$ and $Property$ is typically indirect: The JSON-LD \texttt{@context} maps to schema.org IRIs, which, when resolved as Linked Data, embed their formal definition as RDFa. Extensions may however include such term definitions directly in the RO-Crate.

@@

An important RO-Crate principle is that of being \textbf{self-described} Therefore the serialisation of the RO-Crate into a file should also describe itself in a \href{https://www.researchobject.org/ro-crate/1.1/root-data-entity.html#ro-crate-metadata-file-descriptor}{Metadata File Descriptor}, indicating it is $about$ (describing) the RO-Crate root data entity, and that it $conformsTo$ a particular version of the RO-Crate specification:

@@

Note that although the metadata file necessarily is an \emph{information resource} written to disk or served over the network (as JSON-LD), it is not considered to be a contained \emph{part} of the RO-Crate in the form of a \emph{data entity}, rather it is described only as a \emph{contextual entity}.

In the conceptual model the \emph{RO-Crate Metadata File} can be seen as the top-level node that describes the \emph{RO-Crate Root}, however in the formal model (and the JSON-LD format) the metadata file descriptor is an additional contextual entity that is not affecting the depth-limit of the RO-Crate.

@@

This exposes the first order logic domain of discourse of IRIs, with rational numbers and strings as their corresponding JSON-LD representation. These production rules first grow the graph of $R$ by adding a transitive rule – anything described in $R$ which is related to $o$, means that $o$ is also mentioned by the $ROCrate(R)$. For simplicity this rule is one-way; in theory the graph can also contain free-standing contextual entities that have outgoing relations to data- and contextual entities, but these are proposed to be bound to the root data entity with schema.org relation \href{http://schema.org/mentions}.
